---
title: "Analytical Walkthrough"
subtitle: "From consumption behavior to segment-specific planogram decisions"
description: "A confidentiality-safe replica of behavioral segmentation and controlled retail experimentation using synthetic data."
page-layout: full

format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    toc: true
    toc-depth: 2

execute:
  enabled: true
  echo: true
  warning: false
  message: false
---

::: {.replica-note}

### About this analytical replica

This walkthrough recreates the analytical logic of a real retail experimentation project using synthetic data and independently written code.

The business problem, segmentation logic, operational constraints and measurement approach reflect the work performed. Store identities, assignments, effect sizes and numerical results are generalized for confidentiality.

:::

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(pwr)

set.seed(42)

options(
  dplyr.summarise.inform = FALSE,
  scipen = 999
)

N_STORES <- 1200
N_PLAZAS <- 60
HISTORY_WEEKS <- 60
POST_WEEKS <- 8

segments <- c(
  "Full Flavor",
  "Light",
  "Menthol",
  "NTD"
)

planograms <- c(
  "Control",
  "Planogram A",
  "Planogram B"
)

# Treatment colors: neutral control + two distinct alternatives
assignment_palette <- c(
  "Control" = "#6B7280",
  "Planogram A" = "#D97706",
  "Planogram B" = "#2563EB"
)

# Segment colors reflect the convention used in the original work
segment_palette <- c(
  "Full Flavor" = "#B42318",
  "Light" = "#D4A017",
  "Menthol" = "#2F6BFF",
  "NTD" = "#7A5AF8"
)

treatment_palette <- c(
  "Control" = "#6B7280",
  "Planogram A" = "#D97706",
  "Planogram B" = "#2563EB"
)
```

## The business question

The objective was not to find one national planogram winner.

> **Which planogram should be used for each consumption profile, and is the evidence strong enough to justify rollout?**

The analysis had five steps:

1. identify meaningful consumption profiles;
2. translate those profiles into an operational Plaza segment;
3. compare Control, Planogram A and Planogram B within each segment;
4. estimate incremental impact versus Control;
5. recommend rollout only where the evidence supported it.

The commercial outcome is expressed in **MIOS — millions of sticks**.

---

## 1. Build meaningful consumption profiles

A simple “largest category wins” rule was not useful because **Full Flavor was structurally the largest category in many stores**. That would have classified too much of the market into the same group.

Instead, each store was represented by its complete historical MIOS mix across **Full Flavor, Light, Menthol and NTD — New Taste Dimensions**.

We used four **centroids** to represent the typical category mix of each consumption profile. The important idea is simple: a store is understood by the **shape of its full consumption mix**, not by whichever category happens to rank first.

```{r}
#| label: generate-stores
#| echo: false
#| results: hide

stores <- tibble(
  store_id = sprintf("S%04d", 1:N_STORES),
  plaza_id = sample(
    sprintf("P%02d", 1:N_PLAZAS),
    N_STORES,
    replace = TRUE
  ),
  store_age_weeks = sample(
    35:180,
    N_STORES,
    replace = TRUE
  ),
  latent_profile = sample(
    segments,
    N_STORES,
    replace = TRUE,
    prob = c(0.40, 0.25, 0.20, 0.15)
  )
)

profile_template <- tribble(
  ~latent_profile, ~full_flavor, ~light, ~menthol, ~ntd,
  "Full Flavor",   0.50,         0.24,   0.16,     0.10,
  "Light",         0.38,         0.36,   0.16,     0.10,
  "Menthol",       0.40,         0.18,   0.32,     0.10,
  "NTD",           0.39,         0.17,   0.14,     0.30
)

store_profiles <- stores %>%
  left_join(
    profile_template,
    by = "latent_profile"
  ) %>%
  rowwise() %>%
  mutate(
    full_flavor_obs = max(
      0.01,
      full_flavor + rnorm(1, 0, 0.025)
    ),
    light_obs = max(
      0.01,
      light + rnorm(1, 0, 0.025)
    ),
    menthol_obs = max(
      0.01,
      menthol + rnorm(1, 0, 0.025)
    ),
    ntd_obs = max(
      0.01,
      ntd + rnorm(1, 0, 0.025)
    ),
    total_obs =
      full_flavor_obs +
      light_obs +
      menthol_obs +
      ntd_obs,
    full_flavor_share =
      full_flavor_obs / total_obs,
    light_share =
      light_obs / total_obs,
    menthol_share =
      menthol_obs / total_obs,
    ntd_share =
      ntd_obs / total_obs
  ) %>%
  ungroup()

centroids <- store_profiles %>%
  group_by(latent_profile) %>%
  summarise(
    full_flavor_centroid = mean(full_flavor_share),
    light_centroid = mean(light_share),
    menthol_centroid = mean(menthol_share),
    ntd_centroid = mean(ntd_share),
    .groups = "drop"
  ) %>%
  rename(segment = latent_profile)
```

```{r}
#| label: centroid-profile-chart
#| fig-cap: "Synthetic behavioral centroids used to interpret Plaza consumption profiles"

centroid_plot <- centroids %>%
  rename(
    `Full Flavor` = full_flavor_centroid,
    Light = light_centroid,
    Menthol = menthol_centroid,
    NTD = ntd_centroid
  ) %>%
  pivot_longer(
    cols = c(
      `Full Flavor`,
      Light,
      Menthol,
      NTD
    ),
    names_to = "category",
    values_to = "share"
  )

ggplot(
  centroid_plot,
  aes(
    x = segment,
    y = share,
    fill = category
  )
) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(values = segment_palette) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Each Segment Represents a Different Consumption Pattern",
    subtitle = "The full category mix—not the largest category alone—defines the profile",
    x = NULL,
    y = "Share of historical MIOS",
    fill = "Category"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank()
  )
```

The chart shows why the segmentation is useful: each profile has a different relative mix, even when Full Flavor remains important across the market.

### Make the segmentation executable

Planograms could not realistically be implemented store by store. Execution happened at the **Plaza** level.

For each Plaza, we combined store-level category shares, calculated the **harmonic mean by category**, normalized the four values into one Plaza consumption profile, and matched that profile to the nearest centroid.

```text
Store-level category shares
        ↓
Harmonic mean by category within Plaza
        ↓
Normalized Plaza consumption profile
        ↓
Nearest centroid
        ↓
Operational Plaza segment
```

The harmonic mean makes the Plaza profile more sensitive to **consistency across stores**. A category therefore needs to be broadly represented in the Plaza rather than looking strong because of only a few stores.

> **Store behavior creates the signal; Plaza segmentation makes it operational.**

---

## 2. Run the A/B test within each segment

Only stores with at least **52 weeks of historical data** were included, giving enough history to account for seasonality, underlying store growth and normal variation.

Control Plazas were selected to be **comparable to treated Plazas on historical performance and consumption profile**, so the comparison did not rely on fundamentally different stores.

Within each Plaza segment, we compared:

- **Control:** existing planogram
- **Planogram A:** color-led treatment adapted to that segment
- **Planogram B:** brand-led treatment adapted to that segment

A and B were **treatment families**, not single national layouts. Each had a Full Flavor, Light, Menthol and NTD version.

```{r}
#| label: eligibility-and-assignment
#| echo: false
#| results: hide

eligible_stores <- store_profiles %>%
  filter(store_age_weeks >= 52)

harmonic_mean <- function(x) {
  length(x) / sum(1 / x)
}

plaza_profiles <- eligible_stores %>%
  group_by(plaza_id) %>%
  summarise(
    full_flavor_hm = harmonic_mean(full_flavor_share),
    light_hm = harmonic_mean(light_share),
    menthol_hm = harmonic_mean(menthol_share),
    ntd_hm = harmonic_mean(ntd_share),
    .groups = "drop"
  ) %>%
  mutate(
    total_hm =
      full_flavor_hm +
      light_hm +
      menthol_hm +
      ntd_hm,
    full_flavor_share = full_flavor_hm / total_hm,
    light_share = light_hm / total_hm,
    menthol_share = menthol_hm / total_hm,
    ntd_share = ntd_hm / total_hm
  )

distance_to_centroid <- function(
  ff, li, me, nt,
  centroid_row
) {
  sqrt(
    (ff - centroid_row$full_flavor_centroid)^2 +
    (li - centroid_row$light_centroid)^2 +
    (me - centroid_row$menthol_centroid)^2 +
    (nt - centroid_row$ntd_centroid)^2
  )
}

plaza_segments <- plaza_profiles %>%
  rowwise() %>%
  mutate(
    distance_ff = distance_to_centroid(
      full_flavor_share,
      light_share,
      menthol_share,
      ntd_share,
      centroids %>% filter(segment == "Full Flavor")
    ),
    distance_light = distance_to_centroid(
      full_flavor_share,
      light_share,
      menthol_share,
      ntd_share,
      centroids %>% filter(segment == "Light")
    ),
    distance_menthol = distance_to_centroid(
      full_flavor_share,
      light_share,
      menthol_share,
      ntd_share,
      centroids %>% filter(segment == "Menthol")
    ),
    distance_ntd = distance_to_centroid(
      full_flavor_share,
      light_share,
      menthol_share,
      ntd_share,
      centroids %>% filter(segment == "NTD")
    ),
    plaza_segment = case_when(
      distance_ff == min(
        distance_ff,
        distance_light,
        distance_menthol,
        distance_ntd
      ) ~ "Full Flavor",
      distance_light == min(
        distance_ff,
        distance_light,
        distance_menthol,
        distance_ntd
      ) ~ "Light",
      distance_menthol == min(
        distance_ff,
        distance_light,
        distance_menthol,
        distance_ntd
      ) ~ "Menthol",
      TRUE ~ "NTD"
    )
  ) %>%
  ungroup()

plaza_assignment <- plaza_segments %>%
  group_by(plaza_segment) %>%
  mutate(
    assignment = sample(
      rep(
        c("Control", "Planogram A", "Planogram B"),
        length.out = n()
      )
    )
  ) %>%
  ungroup() %>%
  select(
    plaza_id,
    plaza_segment,
    assignment
  )

eligible_stores <- eligible_stores %>%
  left_join(
    plaza_assignment,
    by = "plaza_id"
  ) %>%
  mutate(
    segment = plaza_segment
  )
```

```{r}
#| label: simulate-panel
#| echo: false
#| results: hide

# Rebuild the analysis table explicitly from the Plaza assignment.
# Dropping possible prior copies makes this robust to duplicated
# assignment/segment columns created by earlier joins.
analysis_stores <- eligible_stores %>%
  select(
    -any_of(
      c(
        "assignment",
        "segment",
        "plaza_segment"
      )
    )
  ) %>%
  left_join(
    plaza_assignment,
    by = "plaza_id"
  ) %>%
  mutate(
    segment = plaza_segment,
    baseline_mios = rlnorm(
      n(),
      meanlog = log(0.09),
      sdlog = 0.30
    ),
    weekly_trend = rnorm(
      n(),
      mean = 0.001,
      sd = 0.0015
    ),
    store_noise = case_when(
      segment == "NTD" ~ runif(
        n(),
        min = 0.10,
        max = 0.20
      ),
      TRUE ~ runif(
        n(),
        min = 0.05,
        max = 0.11
      )
    )
  )

weeks <- tibble(
  week = seq(
    -HISTORY_WEEKS,
    POST_WEEKS - 1
  )
)

weekly_data <- crossing(
  analysis_stores,
  weeks
) %>%
  mutate(
    post = week >= 0,

    seasonal_factor =
      1 +
      0.035 * sin(
        2 * pi * (week + HISTORY_WEEKS) / 52
      ),

    organic_growth =
      1 + weekly_trend * (week + HISTORY_WEEKS),

    treatment_effect = case_when(
      !post ~ 0,

      .data$assignment == "Planogram A" &
        .data$segment == "Full Flavor" ~ 0.060,

      .data$assignment == "Planogram B" &
        .data$segment == "Full Flavor" ~ 0.020,

      .data$assignment == "Planogram A" &
        .data$segment == "Light" ~ 0.015,

      .data$assignment == "Planogram B" &
        .data$segment == "Light" ~ 0.055,

      .data$assignment == "Planogram A" &
        .data$segment == "Menthol" ~ 0.000,

      .data$assignment == "Planogram B" &
        .data$segment == "Menthol" ~ 0.060,

      .data$assignment == "Planogram A" &
        .data$segment == "NTD" ~ 0.075,

      .data$assignment == "Planogram B" &
        .data$segment == "NTD" ~ 0.025,

      TRUE ~ 0
    ),

    mios =
      baseline_mios *
      seasonal_factor *
      organic_growth *
      (1 + treatment_effect) *
      exp(
        rnorm(
          n(),
          mean = 0,
          sd = store_noise
        )
      )
  )
```

```{r}
#| label: segment-trends
#| fig-cap: "Indexed weekly MIOS by segment-specific planogram treatment"

trend_data <- weekly_data %>%
  filter(
    week >= -8,
    week <= 7
  ) %>%
  group_by(
    segment,
    assignment,
    week
  ) %>%
  summarise(
    weekly_mios = mean(mios),
    .groups = "drop"
  ) %>%
  group_by(
    segment,
    assignment
  ) %>%
  mutate(
    pre_period_mean = mean(
      weekly_mios[week < 0]
    ),
    indexed_mios =
      100 * weekly_mios / pre_period_mean
  ) %>%
  ungroup()

ggplot(
  trend_data,
  aes(
    x = week,
    y = indexed_mios,
    color = assignment
  )
) +
  geom_vline(
    xintercept = -0.5,
    linetype = "dashed",
    color = "#9CA3AF"
  ) +
  geom_line(linewidth = 1.05) +
  geom_point(size = 1.8) +
  facet_wrap(
    ~ segment,
    scales = "fixed"
  ) +
  scale_color_manual(values = treatment_palette) +
  labs(
    title = "Matched Groups Follow Similar Pre-Test Trends",
    subtitle = "Weekly MIOS is indexed to each group's pre-period average (= 100) so the chart focuses on relative change after implementation",
    x = "Week relative to implementation",
    y = "Indexed MIOS (pre-period = 100)",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

The chart is indexed to each group's pre-period average so differences in absolute store size do not dominate the visual. What matters is whether matched Control and treatment groups followed similar pre-test patterns and then diverged after implementation.

It is also split by segment on purpose. A national average would hide the question we actually care about: **does the best treatment differ by consumption profile?**

---

## 3. Measure incremental impact versus Control

A before-and-after increase is not enough to attribute the change to the planogram.

```text
Change in treated Plazas
        minus
Change in Control Plazas
        =
Incremental planogram effect
```

This is the intuition behind a **difference-in-differences style comparison**.

In business terms:

> **We want to know whether MIOS increased more than it would have without the planogram change.**

```{r}
#| label: segment-model
#| echo: false
#| results: hide

experiment_window <- weekly_data %>%
  filter(
    week >= -8,
    week <= 7
  ) %>%
  mutate(
    assignment = factor(
      assignment,
      levels = c(
        "Control",
        "Planogram A",
        "Planogram B"
      )
    ),
    period = if_else(
      post,
      "Post",
      "Pre"
    )
  )

plaza_segment_changes <- experiment_window %>%
  group_by(
    plaza_id,
    segment,
    assignment,
    period
  ) %>%
  summarise(
    mean_mios = mean(mios),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = period,
    values_from = mean_mios
  ) %>%
  mutate(
    absolute_change = Post - Pre,
    relative_change = Post / Pre - 1
  )

segment_effects <- plaza_segment_changes %>%
  group_by(segment) %>%
  group_modify(
    ~ {
      model <- lm(
        absolute_change ~ assignment,
        data = .x
      )

      coefs <- summary(model)$coefficients

      get_term <- function(term, field) {
        if (term %in% rownames(coefs)) {
          coefs[term, field]
        } else {
          NA_real_
        }
      }

      tibble(
        planogram = c(
          "Planogram A",
          "Planogram B"
        ),
        estimate = c(
          get_term(
            "assignmentPlanogram A",
            "Estimate"
          ),
          get_term(
            "assignmentPlanogram B",
            "Estimate"
          )
        ),
        std_error = c(
          get_term(
            "assignmentPlanogram A",
            "Std. Error"
          ),
          get_term(
            "assignmentPlanogram B",
            "Std. Error"
          )
        ),
        p_value = c(
          get_term(
            "assignmentPlanogram A",
            "Pr(>|t|)"
          ),
          get_term(
            "assignmentPlanogram B",
            "Pr(>|t|)"
          )
        )
      )
    }
  ) %>%
  ungroup() %>%
  mutate(
    lower_95 = estimate - 1.96 * std_error,
    upper_95 = estimate + 1.96 * std_error,
    statistically_clear =
      !is.na(p_value) &
      p_value < 0.05 &
      estimate > 0
  )
```

```{r}
#| label: segment-effect-plot
#| fig-cap: "Incremental MIOS effects by consumption segment"

ggplot(
  segment_effects,
  aes(
    x = segment,
    y = estimate,
    ymin = lower_95,
    ymax = upper_95,
    color = segment,
    shape = planogram
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "#9CA3AF"
  ) +
  geom_errorbar(
    width = 0.10,
    position = position_dodge(width = 0.45),
    linewidth = 0.8
  ) +
  geom_point(
    position = position_dodge(width = 0.45),
    size = 3
  ) +
  coord_flip() +
  scale_color_manual(values = segment_palette) +
  labs(
    title = "The Recommended Planogram Can Differ by Segment",
    subtitle = "Incremental MIOS versus Control, with uncertainty around each estimate",
    x = NULL,
    y = "Incremental MIOS vs Control",
    color = NULL,
    shape = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank()
  )
```

The important result is not “A beats B nationally.” It is whether **A or B produces a credible incremental effect within each segment**.

---

## 4. Separate visible growth from credible evidence

Some planograms can show attractive percentage growth and still be too variable to support rollout.

That is why the analysis considers both **effect size** and **uncertainty**. A large raw increase is not automatically a reliable treatment effect.

This mattered when communicating results to Marketing and Commercial teams: the recommendation had to explain **why an attractive percentage could still be insufficient evidence for rollout**.

---

## 5. Turn the experiment into a decision

For each consumption segment:

> **Roll out the treatment only when it shows a positive incremental effect versus Control and the evidence is stable enough to support action.**

If neither A nor B clears that bar, keep the current configuration and continue testing.

The management takeaway is:

> **Use the planogram with credible incremental evidence for that consumption profile. Do not force one national winner when different segments support different decisions.**

---

# Technical appendix

## Data-quality checks

```{r}
#| label: validation
#| echo: false

validation_checks <- tibble(
  check = c(
    "Duplicate store-week records",
    "Stores with <52 weeks history included",
    "Stores assigned to multiple Plazas",
    "Plazas assigned to multiple planograms",
    "Missing planogram assignment",
    "Non-positive MIOS values"
  ),
  invalid_records = c(
    weekly_data %>%
      count(store_id, week) %>%
      filter(n > 1) %>%
      nrow(),

    weekly_data %>%
      distinct(store_id, store_age_weeks) %>%
      filter(store_age_weeks < 52) %>%
      nrow(),

    weekly_data %>%
      distinct(store_id, plaza_id) %>%
      count(store_id) %>%
      filter(n > 1) %>%
      nrow(),

    plaza_assignment %>%
      count(plaza_id) %>%
      filter(n > 1) %>%
      nrow(),

    weekly_data %>%
      filter(is.na(assignment)) %>%
      nrow(),

    weekly_data %>%
      filter(mios <= 0) %>%
      nrow()
  )
)

knitr::kable(
  validation_checks,
  col.names = c(
    "Validation check",
    "Invalid records"
  ),
  align = c("l", "r")
)
```

```{r}
#| label: validation-result
#| echo: false
#| results: asis

total_invalid <- sum(
  validation_checks$invalid_records
)

if (total_invalid == 0) {
  cat(
    "**Validation result:** All structural checks passed."
  )
} else {
  cat(
    paste0(
      "**Validation result:** ",
      total_invalid,
      " records require investigation before interpretation."
    )
  )
}
```

## Illustrative SQL

```sql
WITH eligible_stores AS (

    SELECT
        store_id,
        plaza_id,
        store_open_date,
        consumption_segment

    FROM store_dimension

    WHERE store_open_date
        <= DATE_SUB(
            experiment_start_date,
            INTERVAL 52 WEEK
        )
),

assigned_stores AS (

    SELECT
        s.store_id,
        s.plaza_id,
        s.consumption_segment,
        a.experiment_group

    FROM eligible_stores AS s

    INNER JOIN planogram_assignment AS a
        ON s.plaza_id = a.plaza_id
),

weekly_outcomes AS (

    SELECT
        s.store_id,
        s.plaza_id,
        s.consumption_segment,
        s.experiment_group,

        DATE_TRUNC(
            d.calendar_date,
            WEEK
        ) AS week,

        SUM(d.mios) AS weekly_mios

    FROM assigned_stores AS s

    INNER JOIN store_daily_performance AS d
        ON s.store_id = d.store_id

    WHERE d.calendar_date BETWEEN
        DATE_SUB(
            experiment_start_date,
            INTERVAL 8 WEEK
        )
        AND
        DATE_ADD(
            experiment_start_date,
            INTERVAL 8 WEEK
        )

    GROUP BY
        s.store_id,
        s.plaza_id,
        s.consumption_segment,
        s.experiment_group,
        week
)

SELECT
    consumption_segment,
    experiment_group,
    week,
    COUNT(DISTINCT store_id) AS stores,
    SUM(weekly_mios) AS mios

FROM weekly_outcomes

GROUP BY
    consumption_segment,
    experiment_group,
    week

ORDER BY
    consumption_segment,
    experiment_group,
    week;
```

## Limitations

This public replica is intentionally simplified. A production analysis would require stronger treatment-allocation diagnostics, Plaza-level clustered inference, missing-data policies, store openings and closures, market shocks, pricing effects and longer-term persistence measurement.

The numerical findings shown here are synthetic and should not be interpreted as actual company results.

---

## What this walkthrough demonstrates

**Demonstrates:** Behavioral segmentation · A/B testing · Causal inference · Statistical analysis · R · SQL

Together, these methods answer one business question: **which planogram should be used for each consumption profile, and when is the evidence strong enough to justify rollout?**

---

[Back to the executive case study](index.qmd){.btn .btn-primary}

[Back to Case Studies](../../projects.qmd){.btn .btn-outline-primary}
