---
title: "Analytical Walkthrough"
subtitle: "From store-level consumption patterns to segment-specific planogram decisions"
description: "A confidentiality-safe replica of behavioral segmentation, matched controls and controlled retail experimentation using synthetic data."
page-layout: full

format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    toc: true
    toc-depth: 2

execute:
  enabled: true
  echo: true
  warning: false
  message: false
---

::: {.replica-note}

### About this analytical replica

This walkthrough recreates the analytical logic of a real retail experimentation project using synthetic data and independently written code.

The business problem, segmentation logic, operational constraints, matching approach and measurement framework reflect the work performed. Store identities, assignments, treatment effects and numerical results are generalized for confidentiality.

:::

```{r}
#| label: setup
#| include: false

library(tidyverse)

set.seed(42)

options(
  dplyr.summarise.inform = FALSE,
  scipen = 999
)

N_STORES <- 1200
N_PLAZAS <- 60
MIN_HISTORY_WEEKS <- 61
HISTORY_WEEKS <- 61
POST_WEEKS <- 8
NETWORK_STORES <- 24000

segments <- c(
  "Full Flavor",
  "Light",
  "Menthol",
  "NTD"
)

segment_y_order <- rev(segments)

treatment_palette <- c(
  "Control" = "#6B7280",
  "Planogram A" = "#D97706",
  "Planogram B" = "#2563EB"
)

segment_palette <- c(
  "Full Flavor" = "#B42318",
  "Light" = "#D4A017",
  "Menthol" = "#2F6BFF",
  "NTD" = "#7A5AF8"
)
```

## The business question

The objective was to answer:

> **Which planogram should be used for each consumption profile, and is the evidence strong enough to justify rollout?**

The analysis followed four decisions:

1. identify meaningful consumption profiles;
2. translate store-level behavior into an executable **Plaza segment**;
3. compare each treated store with **three controls matched on historical trend and seasonality**;
4. estimate incremental MIOS and recommend the planogram only where the evidence supported it.

The commercial outcome is expressed in **MIOS — millions of sticks**.


---

## What this walkthrough demonstrates

**Demonstrates:** Behavioral segmentation · A/B testing · Matched controls · Nearest-neighbor matching · Causal inference · Statistical analysis · R · SQL

More importantly, it demonstrates a complete decision path:

> **Understand consumer behavior → make the strategy executable → build a credible counterfactual → estimate incremental impact → recommend action only where the evidence supports it.**

---

## 1. Build meaningful consumption profiles

A simple “largest category wins” rule was not useful because **Full Flavor was structurally the largest category in many stores**. That would have pushed too much of the market into one group and hidden meaningful differences in the rest of the consumption mix.

Instead, each store was represented by its complete historical MIOS mix across:

- **Full Flavor**
- **Light**
- **Menthol**
- **NTD — New Taste Dimensions**

### Build four business-defined reference profiles

The four segments were known in advance from the business context. The analytical problem was therefore not to discover an arbitrary number of clusters, but to create a **representative consumption profile for each of the four categories**.

We first calculated the **national average category mix**. Then, for every store, we measured how much each category over- or under-indexed relative to that national pattern.

Stores showing a clear relative over-index toward Full Flavor, Light, Menthol or NTD were used as the anchor group for that category.

For each anchor group, we calculated the average four-category consumption mix. That average became the **reference profile** for the segment.

```text
Historical store mix
        ↓
Compare with national category mix
        ↓
Identify stores with a clear relative over-index
        ↓
Average their full four-category mix
        ↓
Four reference profiles:
Full Flavor · Light · Menthol · NTD
```

This matters because a store can still sell more Full Flavor than any other single category and yet behave more like a **Menthol** or **NTD** store relative to the national pattern.

Stores without a clearly dominant relative pattern — and later the aggregated Plaza profiles — were assigned to the **closest reference profile based on their complete consumption mix**.

```{r}
#| label: generate-segmentation-data
#| echo: false
#| results: hide

plazas <- tibble(
  plaza_id = sprintf("P%02d", 1:N_PLAZAS),
  plaza_profile = rep(
    segments,
    length.out = N_PLAZAS
  )
) %>%
  slice_sample(prop = 1)

stores <- tibble(
  store_id = sprintf("S%04d", 1:N_STORES),
  plaza_id = sample(
    plazas$plaza_id,
    N_STORES,
    replace = TRUE
  ),
  store_age_weeks = sample(
    40:180,
    N_STORES,
    replace = TRUE
  )
) %>%
  left_join(
    plazas,
    by = "plaza_id"
  ) %>%
  rowwise() %>%
  mutate(
    # Most stores resemble the behavioral profile of their Plaza,
    # while some retain realistic within-Plaza heterogeneity.
    latent_profile = if (
      runif(1) < 0.78
    ) {
      plaza_profile
    } else {
      sample(
        segments[
          segments != plaza_profile
        ],
        1
      )
    }
  ) %>%
  ungroup()

# Synthetic population used only for the public replica.
population_template <- tribble(
  ~latent_profile, ~full_flavor, ~light, ~menthol, ~ntd,
  "Full Flavor",   0.50,         0.24,   0.16,     0.10,
  "Light",         0.42,         0.32,   0.16,     0.10,
  "Menthol",       0.43,         0.18,   0.29,     0.10,
  "NTD",           0.43,         0.17,   0.14,     0.26
)

store_profiles <- stores %>%
  left_join(
    population_template,
    by = "latent_profile"
  ) %>%
  rowwise() %>%
  mutate(
    ff_raw = max(0.01, full_flavor + rnorm(1, 0, 0.025)),
    light_raw = max(0.01, light + rnorm(1, 0, 0.025)),
    menthol_raw = max(0.01, menthol + rnorm(1, 0, 0.025)),
    ntd_raw = max(0.01, ntd + rnorm(1, 0, 0.025)),
    total_raw = ff_raw + light_raw + menthol_raw + ntd_raw,
    full_flavor_share = ff_raw / total_raw,
    light_share = light_raw / total_raw,
    menthol_share = menthol_raw / total_raw,
    ntd_share = ntd_raw / total_raw
  ) %>%
  ungroup()

national_mix <- store_profiles %>%
  summarise(
    full_flavor_share = mean(full_flavor_share),
    light_share = mean(light_share),
    menthol_share = mean(menthol_share),
    ntd_share = mean(ntd_share)
  )

store_index <- store_profiles %>%
  mutate(
    ff_index = full_flavor_share / national_mix$full_flavor_share,
    light_index = light_share / national_mix$light_share,
    menthol_index = menthol_share / national_mix$menthol_share,
    ntd_index = ntd_share / national_mix$ntd_share,
    max_index = pmax(
      ff_index,
      light_index,
      menthol_index,
      ntd_index
    ),
    anchor_segment = case_when(
      max_index < 1.12 ~ NA_character_,
      ff_index == max_index ~ "Full Flavor",
      light_index == max_index ~ "Light",
      menthol_index == max_index ~ "Menthol",
      TRUE ~ "NTD"
    )
  )

reference_profiles <- store_index %>%
  filter(!is.na(anchor_segment)) %>%
  group_by(anchor_segment) %>%
  summarise(
    full_flavor = mean(full_flavor_share),
    light = mean(light_share),
    menthol = mean(menthol_share),
    ntd = mean(ntd_share),
    .groups = "drop"
  ) %>%
  rename(segment = anchor_segment)
```

```{r}
#| label: reference-profile-chart
#| fig-cap: "Synthetic reference profiles used to illustrate the segmentation logic"

reference_plot <- reference_profiles %>%
  pivot_longer(
    cols = c(
      full_flavor,
      light,
      menthol,
      ntd
    ),
    names_to = "category",
    values_to = "share"
  ) %>%
  mutate(
    category = recode(
      category,
      full_flavor = "Full Flavor",
      light = "Light",
      menthol = "Menthol",
      ntd = "NTD"
    ),
    category = factor(
      category,
      levels = c(
        "Full Flavor",
        "Light",
        "Menthol",
        "NTD"
      )
    ),
    segment = factor(
      segment,
      levels = segment_y_order
    )
  )

ggplot(
  reference_plot,
  aes(
    x = share,
    y = segment,
    fill = category
  )
) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = segment_palette) +
  scale_x_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    title = "What Defines Each Reference Profile?",
    subtitle = "Each horizontal bar shows the category shares within one reference profile",
    x = "Share of historical MIOS",
    y = NULL,
    fill = "Category"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top",
    axis.text.y = element_text(face = "bold"),
    plot.title.position = "plot"
  )
```

Each horizontal bar sums to 100%, so all profiles have the same total length and can be compared on composition alone. This chart is descriptive: it helps the reader see the shape of each profile. The assignment rule still compares stores with these profiles relative to the national mix, which is why a store can still be classified as Light, Menthol or NTD even when Full Flavor remains the largest absolute share.

### Make the segmentation executable

Planograms could not realistically be implemented store by store. Execution happened at the **Plaza** level.

For each Plaza, store-level category shares were combined using the **harmonic mean by category**. The four resulting values were normalized into one Plaza consumption profile and matched to the nearest behavioral reference profile.

```text
Store-level category shares
        ↓
Harmonic mean by category within Plaza
        ↓
Normalized Plaza consumption profile
        ↓
Nearest reference profile
        ↓
Operational Plaza segment
```

The harmonic mean makes the Plaza profile more sensitive to **consistency across stores**. A category needs to be broadly represented across the Plaza rather than looking strong because of only a few stores. It also keeps the classification focused on the **distribution of category shares across stores**, rather than on absolute sales volume alone. It also keeps the classification focused on the **distribution of category shares across stores**, rather than on absolute sales volume alone.

> **Store behavior creates the signal; Plaza segmentation makes it operational.**

```{r}
#| label: plaza-segmentation
#| echo: false
#| results: hide

eligible_stores <- store_profiles %>%
  filter(store_age_weeks >= MIN_HISTORY_WEEKS)

harmonic_mean <- function(x) {
  length(x) / sum(1 / x)
}

plaza_profiles <- eligible_stores %>%
  group_by(plaza_id) %>%
  summarise(
    full_flavor = harmonic_mean(full_flavor_share),
    light = harmonic_mean(light_share),
    menthol = harmonic_mean(menthol_share),
    ntd = harmonic_mean(ntd_share),
    .groups = "drop"
  ) %>%
  mutate(
    total = full_flavor + light + menthol + ntd,
    full_flavor = full_flavor / total,
    light = light / total,
    menthol = menthol / total,
    ntd = ntd / total
  ) %>%
  select(-total)

reference_matrix <- reference_profiles %>%
  select(
    full_flavor,
    light,
    menthol,
    ntd
  ) %>%
  as.matrix()

nearest_reference <- function(ff, li, me, nt) {
  profile <- c(ff, li, me, nt)

  distances <- rowSums(
    (
      reference_matrix -
      matrix(
        profile,
        nrow = nrow(reference_matrix),
        ncol = 4,
        byrow = TRUE
      )
    )^2
  )

  reference_profiles$segment[which.min(distances)]
}

plaza_segments <- plaza_profiles %>%
  rowwise() %>%
  mutate(
    plaza_segment = nearest_reference(
      full_flavor,
      light,
      menthol,
      ntd
    )
  ) %>%
  ungroup()

missing_plaza_segments <- setdiff(
  segments,
  unique(plaza_segments$plaza_segment)
)

if (length(missing_plaza_segments) > 0) {
  stop(
    paste(
      "Synthetic replica is missing Plaza segments:",
      paste(
        missing_plaza_segments,
        collapse = ", "
      )
    )
  )
}
```

---

## 2. Build a credible comparison with matched controls

Only stores with **more than one year plus eight weeks of history** were eligible. That history was used to capture each store's underlying **trend and seasonal pattern** before the planogram change.

Planogram A and Planogram B were treatment families, each adapted to the Plaza segment:

- **Planogram A:** color-led treatment;
- **Planogram B:** brand-led treatment;
- **Control:** existing configuration, largely organized by price tier, largely organized by price tier.

Each treated store was then matched to **three control stores** with the most similar historical trend and seasonality.

To find the three closest controls efficiently, the original analysis used a **KD-tree nearest-neighbor search** based on historical trend and seasonality.

The public replica reproduces the same nearest-neighbor logic directly in R so the matching approach is easy to inspect without adding an extra package dependency.

```{r}
#| label: treatment-and-panel
#| echo: false
#| results: hide

plaza_assignment <- plaza_segments %>%
  group_by(plaza_segment) %>%
  mutate(
    assignment = sample(
      rep(
        c(
          "Control",
          "Planogram A",
          "Planogram B"
        ),
        length.out = n()
      )
    )
  ) %>%
  ungroup() %>%
  select(
    plaza_id,
    plaza_segment,
    assignment
  )

analysis_stores <- eligible_stores %>%
  left_join(
    plaza_assignment,
    by = "plaza_id"
  ) %>%
  mutate(
    segment = plaza_segment,
    baseline_mios = rlnorm(
      n(),
      meanlog = log(0.09),
      sdlog = 0.28
    ),
    weekly_trend = rnorm(
      n(),
      mean = 0.0010,
      sd = 0.0014
    ),
    seasonal_amplitude = runif(
      n(),
      min = 0.015,
      max = 0.055
    ),
    seasonal_phase = runif(
      n(),
      min = -pi,
      max = pi
    ),
    store_noise = case_when(
      segment == "NTD" &
        assignment == "Planogram B" ~ runif(n(), 0.30, 0.48),
      segment == "NTD" ~ runif(n(), 0.055, 0.085),
      TRUE ~ runif(n(), 0.045, 0.09)
    )
  )

weeks <- tibble(
  week = seq(
    -HISTORY_WEEKS,
    POST_WEEKS - 1
  )
)

weekly_data <- crossing(
  analysis_stores,
  weeks
) %>%
  mutate(
    post = week >= 0,
    seasonal_factor =
      1 +
      seasonal_amplitude *
      sin(
        2 * pi * (week + HISTORY_WEEKS) / 52 +
        seasonal_phase
      ),
    organic_growth =
      1 + weekly_trend * (week + HISTORY_WEEKS),
    treatment_effect = case_when(
      !post ~ 0,

      assignment == "Planogram A" &
        segment == "Full Flavor" ~ 0.060,

      assignment == "Planogram B" &
        segment == "Full Flavor" ~ -0.005,

      assignment == "Planogram A" &
        segment == "Light" ~ 0.000,

      assignment == "Planogram B" &
        segment == "Light" ~ 0.055,

      assignment == "Planogram A" &
        segment == "Menthol" ~ -0.010,

      assignment == "Planogram B" &
        segment == "Menthol" ~ 0.060,

      assignment == "Planogram A" &
        segment == "NTD" ~ 0.030,

      assignment == "Planogram B" &
        segment == "NTD" ~ 0.070,

      TRUE ~ 0
    ),
    mios =
      baseline_mios *
      seasonal_factor *
      organic_growth *
      (1 + treatment_effect) *
      exp(
        rnorm(
          n(),
          mean = 0,
          sd = store_noise
        )
      )
  )

matching_features <- analysis_stores %>%
  group_by(segment) %>%
  mutate(
    trend_z =
      (weekly_trend - mean(weekly_trend)) /
      sd(weekly_trend),
    season_z =
      (seasonal_amplitude - mean(seasonal_amplitude)) /
      sd(seasonal_amplitude)
  ) %>%
  ungroup()

treated_stores <- matching_features %>%
  filter(assignment != "Control")

control_pool <- matching_features %>%
  filter(assignment == "Control")

matched_pairs <- map_dfr(
  seq_len(nrow(treated_stores)),
  function(i) {

    treated <- treated_stores[i, ]

    candidates <- control_pool %>%
      filter(segment == treated$segment) %>%
      mutate(
        distance =
          (trend_z - treated$trend_z)^2 +
          (season_z - treated$season_z)^2
      ) %>%
      arrange(distance) %>%
      slice_head(n = 3)

    tibble(
      treated_store_id = treated$store_id,
      control_store_id = candidates$store_id,
      segment = treated$segment,
      planogram = treated$assignment,
      control_rank = seq_len(nrow(candidates))
    )
  }
)

matched_segment_check <- matched_pairs %>%
  distinct(segment) %>%
  pull(segment)

missing_matched_segments <- setdiff(
  segments,
  matched_segment_check
)

if (length(missing_matched_segments) > 0) {
  stop(
    paste(
      "Synthetic matching is missing segments:",
      paste(
        missing_matched_segments,
        collapse = ", "
      )
    )
  )
}
```

### Did the matching create comparable histories?

The most useful diagnostic is not whether treated and control stores had the same absolute MIOS, because stores can differ in size.

What matters for this matching strategy is whether their **historical trajectories and seasonal patterns moved similarly before treatment**.

```{r}
#| label: matching-chart
#| fig-cap: "Pre-test trajectories for treated stores and their matched controls"
#| fig-height: 10.5
#| fig-width: 11

treated_pre <- weekly_data %>%
  filter(
    week >= -8,
    week <= -1,
    assignment != "Control"
  ) %>%
  transmute(
    treated_store_id = store_id,
    segment,
    week,
    role = "Treated",
    weekly_mios = mios
  ) %>%
  semi_join(
    matched_pairs %>%
      distinct(treated_store_id),
    by = "treated_store_id"
  )

control_pre <- matched_pairs %>%
  left_join(
    weekly_data %>%
      filter(
        week >= -8,
        week <= -1
      ) %>%
      select(
        store_id,
        week,
        mios
      ),
    by = c(
      "control_store_id" = "store_id"
    )
  ) %>%
  group_by(
    treated_store_id,
    segment,
    week
  ) %>%
  summarise(
    weekly_mios = mean(mios),
    .groups = "drop"
  ) %>%
  mutate(role = "Matched controls")

matching_plot_data <- bind_rows(
  treated_pre,
  control_pre
) %>%
  mutate(
    segment = factor(
      segment,
      levels = segments
    )
  ) %>%
  group_by(
    treated_store_id,
    segment,
    role
  ) %>%
  mutate(
    indexed_mios =
      100 * weekly_mios / mean(weekly_mios)
  ) %>%
  ungroup() %>%
  group_by(
    segment,
    role,
    week
  ) %>%
  summarise(
    indexed_mios = mean(indexed_mios),
    .groups = "drop"
  )

# Defensive check: every segment/role must contain all eight pre-test weeks.
matching_completeness <- matching_plot_data %>%
  count(
    segment,
    role,
    name = "weeks_observed"
  )

if (any(matching_completeness$weeks_observed != 8)) {
  stop(
    "The matching diagnostic is missing one or more pre-test weeks."
  )
}

# Use a shared Y range that contains every observed point.
# This avoids clipping values and visually breaking otherwise complete lines.
y_min <- floor(
  min(matching_plot_data$indexed_mios) - 0.25
)

y_max <- ceiling(
  max(matching_plot_data$indexed_mios) + 0.25
)

ggplot(
  matching_plot_data,
  aes(
    x = week,
    y = indexed_mios,
    color = role,
    group = role
  )
) +
  geom_hline(
    yintercept = 100,
    linetype = "dotted",
    color = "#B8C0CC",
    linewidth = 0.7
  ) +
  geom_line(linewidth = 0.95) +
  geom_point(size = 2.1) +
  facet_wrap(
    ~ segment,
    ncol = 1
  ) +
  scale_color_manual(
    values = c(
      "Treated" = "#2563EB",
      "Matched controls" = "#6B7280"
    )
  ) +
  scale_x_continuous(
    breaks = -8:-1,
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    breaks = seq(
      y_min,
      y_max,
      by = 1
    ),
    expand = expansion(mult = c(0.03, 0.03))
  ) +
  coord_cartesian(
    ylim = c(
      y_min,
      y_max
    )
  ) +
  labs(
    title = "Matched Controls Follow Similar Pre-Test Patterns",
    subtitle = "Before implementation, treated stores and their three matched controls show similar indexed MIOS trajectories across all four segments",
    x = "Week before implementation",
    y = "Indexed MIOS (pre-period average = 100)",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(
      color = "#E5E7EB",
      linewidth = 0.5
    ),
    strip.text = element_text(
      face = "bold",
      size = 12,
      margin = margin(t = 8, b = 5)
    ),
    panel.spacing.y = unit(0.8, "lines"),
    axis.title.x = element_text(
      margin = margin(t = 10)
    ),
    axis.title.y = element_text(
      margin = margin(r = 10)
    ),
    plot.title.position = "plot",
    plot.margin = margin(8, 12, 8, 8)
  )
```

Before implementation, treated stores and their matched controls follow similar historical patterns across all four segments. This supports using the matched controls as a credible comparison for estimating what would likely have happened without the planogram change.

---

## 3. Measure incremental impact versus matched controls

A before-and-after change within a store is not enough.

For every treated store, we compare its change with the average change across its three matched controls:

```text
Treated store: post − pre
        minus
3 matched controls: post − pre
        =
Incremental planogram effect
```

This is the business intuition behind a matched-control, difference-in-differences style comparison.

> **The question is whether MIOS increased more than it likely would have without the planogram change.**

```{r}
#| label: matched-effect-model
#| echo: false
#| results: hide

store_period <- weekly_data %>%
  filter(
    week >= -8,
    week <= 7
  ) %>%
  mutate(
    period = if_else(
      week < 0,
      "Pre",
      "Post"
    )
  ) %>%
  group_by(
    store_id,
    period
  ) %>%
  summarise(
    mean_mios = mean(mios),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = period,
    values_from = mean_mios
  ) %>%
  mutate(
    change = Post - Pre
  )

treated_changes <- treated_stores %>%
  select(
    treated_store_id = store_id,
    segment,
    planogram = assignment
  ) %>%
  left_join(
    store_period %>%
      select(
        treated_store_id = store_id,
        treated_change = change
      ),
    by = "treated_store_id"
  )

matched_control_changes <- matched_pairs %>%
  left_join(
    store_period %>%
      select(
        control_store_id = store_id,
        control_change = change
      ),
    by = "control_store_id"
  ) %>%
  group_by(
    treated_store_id
  ) %>%
  summarise(
    matched_control_change =
      mean(control_change),
    .groups = "drop"
  )

matched_effects <- treated_changes %>%
  left_join(
    matched_control_changes,
    by = "treated_store_id"
  ) %>%
  mutate(
    incremental_mios =
      treated_change -
      matched_control_change
  )

segment_effects <- matched_effects %>%
  group_by(
    segment,
    planogram
  ) %>%
  summarise(
    stores = n(),
    estimate = mean(incremental_mios),
    std_error =
      sd(incremental_mios) / sqrt(n()),
    lower_95 =
      estimate - 1.96 * std_error,
    upper_95 =
      estimate + 1.96 * std_error,
    .groups = "drop"
  )

# Estimate the number of stores represented by each segment at
# an approximately 24K-store network scale.
network_segment_sizes <- analysis_stores %>%
  count(
    segment,
    name = "synthetic_stores"
  ) %>%
  mutate(
    network_stores =
      round(
        synthetic_stores /
          sum(synthetic_stores) *
          NETWORK_STORES
      )
  ) %>%
  select(
    segment,
    network_stores
  )

segment_effects <- segment_effects %>%
  left_join(
    network_segment_sizes,
    by = "segment"
  ) %>%
  mutate(
    # Synthetic NTD example deliberately illustrates an important
    # decision pattern from the original work:
    # the larger observed lift is not necessarily the more reliable one.
    estimate = case_when(
      segment == "Full Flavor" &
        planogram == "Planogram B" ~ -0.0015,
      segment == "NTD" &
        planogram == "Planogram A" ~ 0.0020,
      segment == "NTD" &
        planogram == "Planogram B" ~ 0.0045,
      TRUE ~ estimate
    ),
    lower_95 = case_when(
      segment == "Full Flavor" &
        planogram == "Planogram B" ~ -0.0024,
      segment == "NTD" &
        planogram == "Planogram A" ~ 0.0008,
      segment == "NTD" &
        planogram == "Planogram B" ~ -0.0015,
      TRUE ~ lower_95
    ),
    upper_95 = case_when(
      segment == "Full Flavor" &
        planogram == "Planogram B" ~ -0.0006,
      segment == "NTD" &
        planogram == "Planogram A" ~ 0.0032,
      segment == "NTD" &
        planogram == "Planogram B" ~ 0.0105,
      TRUE ~ upper_95
    ),
    network_estimate =
      estimate * network_stores,
    network_lower_95 =
      lower_95 * network_stores,
    network_upper_95 =
      upper_95 * network_stores
  )
```

```{r}
#| label: effect-chart
#| fig-cap: "Synthetic incremental weekly MIOS versus matched controls"
#| fig-height: 6.8
#| fig-width: 11

effect_plot_data <- segment_effects %>%
  mutate(
    segment = factor(
      segment,
      levels = segment_y_order
    ),
    planogram = factor(
      planogram,
      levels = c(
        "Planogram A",
        "Planogram B"
      )
    ),
    evidence = factor(
      case_when(
        network_lower_95 > 0 ~ "Conclusive positive",
        network_upper_95 < 0 ~ "Conclusive negative",
        TRUE ~ "Inconclusive"
      ),
      levels = c(
        "Conclusive negative",
        "Conclusive positive",
        "Inconclusive"
      )
    )
  )

x_min <- min(effect_plot_data$network_lower_95)
x_max <- max(effect_plot_data$network_upper_95)
x_pad <- max(
  (x_max - x_min) * 0.10,
  2
)

ggplot(
  effect_plot_data,
  aes(
    x = network_estimate,
    y = segment,
    xmin = network_lower_95,
    xmax = network_upper_95,
    color = evidence,
    shape = planogram,
    group = planogram
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "#9CA3AF",
    linewidth = 0.8
  ) +
  geom_errorbarh(
    height = 0,
    linewidth = 1.15,
    position = position_dodge(width = 0.34),
    show.legend = TRUE
  ) +
  geom_point(
    size = 5.2,
    position = position_dodge(width = 0.34)
  ) +
  scale_color_manual(
    values = c(
      "Conclusive positive" = "#16A34A",
      "Inconclusive" = "#EAB308",
      "Conclusive negative" = "#DC2626"
    )
  ) +
  scale_shape_manual(
    values = c(
      "Planogram A" = 16,
      "Planogram B" = 17
    ),
    labels = c(
      "Planogram A (circle)",
      "Planogram B (triangle)"
    )
  ) +
  scale_x_continuous(
    labels = scales::label_number(
      accuracy = 1
    ),
    limits = c(
      x_min - x_pad,
      x_max + x_pad
    )
  ) +
  scale_y_discrete(drop = FALSE) +
  guides(
    color = guide_legend(
      order = 1,
      nrow = 1,
      override.aes = list(
        shape = NA,
        linewidth = 1.6
      )
    ),
    shape = guide_legend(
      order = 2,
      nrow = 1,
      override.aes = list(
        color = "black",
        size = 5.4,
        linewidth = 0
      )
    )
  ) +
  labs(
    title = "Incremental Weekly MIOS by Segment and Planogram",
    subtitle = "Circle = A; triangle = B. Green = positive, yellow = inconclusive, red = negative versus matched controls.",
    x = "Incremental weekly MIOS across the network",
    y = NULL,
    color = NULL,
    shape = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.box = "vertical",
    legend.direction = "horizontal",
    legend.margin = margin(0, 0, 4, 0),
    legend.spacing.y = unit(2, "pt"),
    plot.margin = margin(8, 28, 8, 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(
      size = 10.5,
      face = "bold"
    ),
    plot.title.position = "plot"
  )
```

Read the chart row by row:

- **0** means the tested planogram is not different from its matched controls;
- values to the **right of 0** indicate positive incremental weekly MIOS, while values to the **left of 0** indicate underperformance;
- the **circle** is Planogram A and the **triangle** is Planogram B;
- **green** means the full interval is above 0, **yellow** means the interval crosses 0, and **red** means the full interval is below 0.

The NTD example highlights an important decision principle: **Planogram A has the smaller estimated lift but a conclusive positive effect, while Planogram B shows a larger observed lift with enough variability that the result remains inconclusive.**

**Not every new planogram beats Control, and the largest observed lift is not automatically the best decision.** The experiment is designed to identify which planogram, if any, has enough incremental evidence to justify rollout within each consumption segment.

---

## 4. Turn statistical evidence into a commercial decision

Some planograms can show attractive percentage growth and still be too variable to justify rollout — as illustrated by **Planogram B for NTD** in the chart above.

That distinction mattered in conversations with Marketing and Commercial teams: **large observed growth is not automatically strong evidence**.

For each segment, the decision rule is:

> **Roll out a new planogram only when its incremental effect versus matched controls is positive and the evidence is precise enough to support action.**

Not every new planogram outperformed Control. If a treatment was inconclusive or underperformed its matched controls, it was **not** recommended. If neither A nor B cleared the bar, the decision was to **keep Control** and continue testing.

The management takeaway is:

> **Use the planogram with credible incremental evidence for that consumption profile.**

---

# Technical appendix

## Why three matched controls?

Using three controls gives each treated store a small comparison set rather than relying on a single potentially idiosyncratic store. The matched-control outcome is the average of those three historical comparators.

## Why a KD-tree?

The business logic is nearest-neighbor matching: find stores with the most similar historical trend and seasonality. A KD-tree makes that search efficient when the candidate control population is large.

The public code above uses the same distance logic directly for transparency. The original workflow used a KD-tree implementation.


## Illustrative SQL

```sql
WITH eligible_stores AS (

    SELECT
        store_id,
        plaza_id,
        store_open_date,
        consumption_segment

    FROM store_dimension

    WHERE store_open_date
        <= DATE_SUB(
            experiment_start_date,
            INTERVAL 61 WEEK
        )
),

weekly_history AS (

    SELECT
        store_id,
        DATE_TRUNC(
            calendar_date,
            WEEK
        ) AS week,
        SUM(mios) AS weekly_mios

    FROM store_daily_performance

    GROUP BY
        store_id,
        week
),

weekly_experiment AS (

    SELECT
        e.store_id,
        e.plaza_id,
        e.consumption_segment,
        a.experiment_group,
        h.week,
        h.weekly_mios

    FROM eligible_stores AS e

    INNER JOIN planogram_assignment AS a
        ON e.plaza_id = a.plaza_id

    INNER JOIN weekly_history AS h
        ON e.store_id = h.store_id

)

SELECT
    consumption_segment,
    experiment_group,
    week,
    COUNT(DISTINCT store_id) AS stores,
    SUM(weekly_mios) AS mios

FROM weekly_experiment

GROUP BY
    consumption_segment,
    experiment_group,
    week

ORDER BY
    consumption_segment,
    experiment_group,
    week;
```

## Limitations

This public replica is intentionally simplified.

Treatment assignment, store identities, segment distributions, effect sizes and numerical results are synthetic. The public matching code recreates the logic of the original KD-tree workflow without reproducing proprietary implementation details.

A production analysis would also account for operational exclusions, missing data, store openings and closures, market shocks, pricing effects, repeated use of control stores and uncertainty at the appropriate implementation level.

---

[Back to the executive case study](index.qmd){.btn .btn-primary}

[Back to Case Studies](../../projects.qmd){.btn .btn-outline-primary}
